---
title: "Data"
output: 
  html_document:
    toc: true
    toc_float: true
---

The PRAM dataset stands for the Pregnancy Risk Assessment Monitoring System. It collects state-specific data on maternal behaviors/experiences before, during, and after birth.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

#### Indicator: 2007 Data-set, Reading in Substance Abuse & No Contraception

```{r}
alcohol_data_2007 = read_csv("./data/PRAM_2007_alcohol.csv")

tobacco_data_2007 = read_csv("./data/PRAM_2007_tobacco.csv")

no_contraception_data_2007 = read_csv("./data/PRAM_2007_no_contraception.csv")

infant_mortality_df = read_csv("./data/PRAM_2007_infantmortality.csv")
```

Cleaning Alcohol data 

```{r}
cleaned_alc_2007 <- alcohol_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -data_value_type) |>
  filter(response != "DRINKER WHO QUIT") |>
  filter(response != "NONDRINKER") |>
  filter( response != "NO") |>
  drop_na(response,geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  # Convert to numeric and remove parentheses
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  # Convert to numeric and remove parentheses


view(cleaned_alc_2007)
```

Cleaning Tobacco Data

```{r}
cleaned_tobac_2007 <- tobacco_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_type) |>
  filter(response != "SMOKER WHO QUIT") |>
  filter(response != "NONSMOKER") |>
  filter(response != "None (0 cig)") |>
  filter( response != "NO") |>
  drop_na(response) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE)

view(cleaned_tobac_2007)
```

#### Alcohol

- `year`  refers to the year data was recorded
- `class` refers to ....
- `topic` refers to the topic of the data
- `question` the question that was recorded 
- `data_source` where the recorded data came from
- `response` participants response to the question 
- `data_value` 
- `low_confidence_limit` the lower bound of the confidence interval 
- `high_confidence-limit` the upper bound of the confidence interval
- `sample_size`
- `break_out`
- `break_out_category`
- `class_id`
- `question_id`
- `location_id`
- `break_out_id`
- `break_out_category_id`
- `response_id`

#### 2007 Data-set: Reading in No Substance Abuse & Contraception


```{r}
no_alcohol_data_2007 = read_csv("./data/PRAM_2007_no_alcohol.csv")

no_tobacco_data_2007 = read_csv("./data/PRAM_2007_no_tobacco.csv")

contraception_data_2007 = read_csv("./data/PRAM_2007_contraception.csv")
```

Cleaning the No Alcohol Data

```{r}
cleaned_no_alc_2007 <- no_alcohol_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  drop_na(response)

view(cleaned_no_alc_2007)
```

Cleaning the No Tobacco Data

```{r}
cleaned_no_tobacco_2007 <- no_tobacco_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  drop_na(response)

view(cleaned_no_tobacco_2007)
```

#### Outcome Data: Cleaning Infant Mortality Rate Data

```{r}

cleaned_infant_mortality <- infant_mortality_df |>
  janitor::clean_names() |>
  select(-data_value_std_err, -data_value_type, -data_value_unit, -data_value_footnote_symbol, -data_value_footnote) |>
  drop_na(response) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE)

view(cleaned_infant_mortality)

```

#### Other aspects of data we are interested in 

Use of Contraception

```{r}
cleaned_contraception_2007 <- contraception_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  filter(response != "YES (CHECKED)") |>
  filter(response != "YES") |>
  drop_na(response)

view(cleaned_contraception_2007)
```

No Contraception Use

```{r}
cleaned_no_contra_2007 <- no_contraception_data_2007 %>%
  janitor::clean_names() %>%
  select(-geolocation, -data_value_type) %>%
  drop_na(response)

view(cleaned_no_contra_2007)
```



