---
title: "Infant Fatality Across the US"
output: 
  html_document:
    toc: true
    toc_float: true
---


```{r, message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(plotly)
library(flexdashboard)
library(dplyr)
library(tidyverse)
library(leaflet)
library(knitr)

alcohol_data_2007 = read_csv("./data/PRAM_2007_alcohol.csv")

tobacco_data_2007 = read_csv("./data/PRAM_2007_tobacco.csv")

no_contraception_data_2007 = read_csv("./data/PRAM_2007_no_contraception.csv")

infant_mortality_df = read_csv("./data/PRAM_2007_infantmortality.csv")

maternal_race = read_csv("./data/PRAM_2007_Maternal_Race.csv")

# cleaned alcohol data 
cleaned_alc_2007 <- alcohol_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -data_value_type) |>
  filter(response != "DRINKER WHO QUIT") |>
  filter(response != "NONDRINKER") |>
  filter( response != "NO") |>
  drop_na(response,geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  # Convert to numeric and remove parentheses
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  # Convert to numeric and remove parentheses

# cleaned tobacco data 

cleaned_tobac_2007 <- tobacco_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_type) |>
  filter(response != "SMOKER WHO QUIT") |>
  filter(response != "NONSMOKER") |>
  filter(response != "None (0 cig)") |>
  filter( response != "NO") |>
  drop_na(response, geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  # Convert to numeric and remove parentheses
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  # Convert to numeric and remove parentheses

cleaned_mat_race <- maternal_race |>
  janitor::clean_names() |>
  select(-data_value_std_err, -data_value_type) |>
  drop_na(response,geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  # Convert to numeric and remove parentheses
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  # Convert to numeric and remove parentheses
  
no_alcohol_data_2007 = read_csv("./data/PRAM_2007_no_alcohol.csv")

no_tobacco_data_2007 = read_csv("./data/PRAM_2007_no_tobacco.csv")

contraception_data_2007 = read_csv("./data/PRAM_2007_contraception.csv")

# cleaned no alcohol data 

cleaned_no_alc_2007 <- no_alcohol_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  drop_na(response)

view(cleaned_no_alc_2007)

# cleaned no tobacco data 

cleaned_no_tobacco_2007 <- no_tobacco_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  drop_na(response)

# cleaned infant mortality 

cleaned_infant_mortality <- infant_mortality_df |>
  janitor::clean_names() |>
  select(-data_value_std_err, -data_value_type, -data_value_unit, -data_value_footnote_symbol, -data_value_footnote) |>
  drop_na(response, geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  # Convert to numeric and remove parentheses
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  # Convert to numeric and remove parentheses

# cleaned conception

cleaned_contraception_2007 <- contraception_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  filter(response != "YES (CHECKED)") |>
  filter(response != "YES") |>
  drop_na(response)

# cleaned non conception

cleaned_no_contra_2007 <- no_contraception_data_2007 %>%
  janitor::clean_names() %>%
  select(-data_value_type) %>%
  drop_na(response) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  # Convert to numeric and remove parentheses
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  # Convert to numeric and remove parentheses

```

Both substance abuse and maternal/infant mortality run rampant in the United States, and with the the use of the CDC 2007 PRAM Data set, we are interested in looking at the relationships between these variables. 

## Data Cleaning

We loaded in data from the PRAMS 2007 dataset. This dataset provided information about many different aspects about maternal health and birth that we were interested in addressing. 

After loading in the data, we cleaned each dataset by removing unecessary columns. We then filtered the responses to tailor the dataframe to each dataset, such as removing "no" responses to if mothers were smoking during conception to create a dataset displaying data for mothers who were smoking during conception. 

We used the seperate function for the "geolocation" variable by splitting it into a longitude and latitude variable that was used in creating our leaflet plots.

## Initial questions

##### Our initial research question
Is there a significant relationship between infant mortality rate and the consumption of Alcohol and Tobacco consumption?

##### How did these questions evolve over the course of the project?
Upon reviewing reviewing the 2007 CDC PRAM data set, and thoroughly understanding the data, we picked our variables of interest that we wanted to test. The data provides robust information on several important factors related to maternal and infant health outcomes. 

##### What new questions did you consider in the course of your analysis?
Other parts of the data that interested us was also the use of contraception in the sample demographic. 


## Motivation for this Project
We are interested in looking at how pregnancy outcomes changes among mothers who drank or smoked cigarettes before conception, and comparing it to mothers who did not abuse any substances. As a group, we were collectively interested in both Maternal Child Health and Substance Abuse and this was an ideal way to combine those two interests. The demographics, social dynamics, and healthcare system in the United States provide a unique perspective on the outcomes of maternal health and substance abuse. We are interested in learning more about the relationship between substance abuse and infant mortality rates. 


## Related Work

##### Alcohol use during pregnancy

The negative impact of smoking during pregnancy is evident in its direct association with adverse birth outcomes. This habit has been linked to a range of complications, including damage to the umbilical cord structure, an elevated risk of miscarriage and ectopic pregnancy, as well as an increased likelihood of low birth weight, placental abruption, preterm birth, and heightened infant mortality rates. 
[Reference](https://pubmed.ncbi.nlm.nih.gov/17884310/)

According to Wolfe et al., within a dataset comprising 4,536,701 birth records, 1.20% exhibited diagnostic codes related to drug or alcohol discharge (n = 54,290). The unadjusted Relative Risks (RRs) for maternal (RR = 2.7), fetal (RR = 1.3), neonatal (RR = 2.4), and post-neonatal (RR = 4.3) mortality were elevated among births diagnosed with drug or alcohol-related conditions. 
[Reference](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3349286/)

##### Tobacco use during pregnancy

The consumption of alcohol during pregnancy has been correlated with a spectrum of adverse birth outcomes. This includes heightened risks of miscarriage, stillbirth, infant mortality, congenital anomalies, low birthweight, reduced gestational age, preterm delivery, and infants being born small for their gestational age.
[Reference](https://pubmed.ncbi.nlm.nih.gov/19797967/)

While specific data on infant mortality rates were not explicitly provided, it is noteworthy that maternal smoking during pregnancy constitutes a well-established risk factor for sudden unexpected infant death (SUID). The risk of SUID more than doubled (adjusted odds ratio [aOR] = 2.44; 95% confidence interval [CI] 2.31–2.57) in the presence of any maternal smoking during pregnancy. Furthermore, a twofold increase in SUID risk was observed when comparing no smoking to smoking just 1 cigarette daily throughout pregnancy. 
[Reference](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6564075/)

# Exploratory Analysis

### Infant Deaths Across the US

```{r}
infant_deaths <- cleaned_infant_mortality %>%
  filter(question == "Indicator of infant currently alive" & response == "NO") %>%
  group_by(location_desc) %>%
  summarize(total_infant_deaths = n())

# Display the table using knitr::kable()
knitr::kable(infant_deaths)
```

The presented table summarizes total infant deaths by state, where each row represents a distinct geographical location. The 'location_desc' column specifies the state, while the 'total_infant_deaths' column reveals the corresponding count of infant deaths in each location. The data indicates variations in infant mortality rates across regions, with some areas reporting higher or lower rates than others. Notably, states like Pennsylvania exhibit lower counts of infant deaths, whereas Alaska and Arkansas show higher counts. Nevertheless, a majority of the data falls within the 35 to 50 range. This summary offers a comprehensive overview of the distribution of infant deaths across diverse geographical locations.

### Maternal Alcohol Use 

```{r}
# Plot of question and responses for alcohol

Maternal_alc_use <- cleaned_alc_2007 |>
  ggplot(aes(x = question, fill = response)) +
  geom_bar(position = "dodge") +
  labs(title = "Questions and Responses", x = "Questions", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 4))+
  labs(
    x = "Question",
    y = "Response",
    title =  "Questions vs Response of Alcohol Consumption"
  )
```

the `Maternal_alc_use` plot shows the responses to the questions asked. From this plot, we can see that from the 3 months before pregnancy and the last 3 months of pregnancy, about 1300 monthers continued drinking the same or more amount of alcohol, 1200 reduced how much they consumed alcohol, and about 700 nondrinkers resumed drinking. It is obvious that more mothers continued drinking the same amount or more alcohol than reducing how much they consumed.

# Discussion
