---
title: "Visualization and Analysis"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(plotly)
library(flexdashboard)
library(dplyr)
library(tidyverse)
library(leaflet)
library(knitr)
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
alcohol_data_2007 = read_csv("./data/PRAM_2007_alcohol.csv")

tobacco_data_2007 = read_csv("./data/PRAM_2007_tobacco.csv")

no_contraception_data_2007 = read_csv("./data/PRAM_2007_no_contraception.csv")

infant_mortality_df = read_csv("./data/PRAM_2007_infantmortality.csv")

maternal_race = read_csv("./data/PRAM_2007_Maternal_Race.csv")

# cleaned alcohol data 
cleaned_alc_2007 <- alcohol_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -data_value_type) |>
  filter(response != "DRINKER WHO QUIT") |>
  filter(response != "NONDRINKER") |>
  filter( response != "NO") |>
  drop_na(response,geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  # Convert to numeric and remove parentheses
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  # Convert to numeric and remove parentheses

# cleaned tobacco data 

cleaned_tobac_2007 <- tobacco_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_type) |>
  filter(response != "SMOKER WHO QUIT") |>
  filter(response != "NONSMOKER") |>
  filter(response != "None (0 cig)") |>
  filter( response != "NO") |>
  drop_na(response, geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  

cleaned_mat_race <- maternal_race |>
  janitor::clean_names() |>
  select(-data_value_std_err, -data_value_type) |>
  drop_na(response,geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  
  
no_alcohol_data_2007 = read_csv("./data/PRAM_2007_no_alcohol.csv")

no_tobacco_data_2007 = read_csv("./data/PRAM_2007_no_tobacco.csv")

contraception_data_2007 = read_csv("./data/PRAM_2007_contraception.csv")

# cleaned no alcohol data 

cleaned_no_alc_2007 <- no_alcohol_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  drop_na(response)

view(cleaned_no_alc_2007)

# cleaned no tobacco data 

cleaned_no_tobacco_2007 <- no_tobacco_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  drop_na(response)

# cleaned infant mortality 

cleaned_infant_mortality <- infant_mortality_df |>
  janitor::clean_names() |>
  select(-data_value_std_err, -data_value_type, -data_value_unit, -data_value_footnote_symbol, -data_value_footnote) |>
  drop_na(response, geolocation) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  

# cleaned conception

cleaned_contraception_2007 <- contraception_data_2007 |>
  janitor::clean_names() |>
  select(-data_value_std_err, -geolocation, -data_value_type) |>
  filter(response != "YES (CHECKED)") |>
  filter(response != "YES") |>
  drop_na(response)

# cleaned non conception

cleaned_no_contra_2007 <- no_contraception_data_2007 %>%
  janitor::clean_names() %>%
  select(-data_value_type) %>%
  drop_na(response) |>
  separate(geolocation, into = c("latitude", "longitude"), sep = ", ", convert = TRUE) |>
   mutate(latitude = as.numeric(str_replace_all(latitude, "\\(|\\)", "")),  
         longitude = as.numeric(str_replace_all(longitude, "\\(|\\)", "")))  

```

We aim to see the patterns of infant mortality among maternal populations based on their tobacco and alcohol consumption patterns. Additionally, we also want to see how demographic factors such as race, age, income levels, educational levels and intent to have children can also contribute to infant mortality rates. 



### Plots of Maternal Alcohol Consumption

```{r, message=FALSE, warning=FALSE}


# Plot of question and responses for alcohol

# Create ggplot object
gg_plot <- cleaned_alc_2007 %>%
  ggplot(aes(x = question, fill = response)) +
  geom_bar(position = "dodge") +
  labs(title = "Questions and Responses", x = "Questions", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 5, size = 2)) +
  labs(
    x = "Question",
    y = "Response",
    title = "Questions vs Response of Alcohol Consumption"
  )

print(gg_plot)

```

```{r}
# Extract data directly from the original data frame
plot_data <- cleaned_alc_2007 %>%
  group_by(question, response) %>%
  summarize(count = n())

# Convert data to Plotly
plot_ly(data = plot_data, x = ~question, y = ~count, color = ~response, type = "bar", split = ~response) %>%
  layout(
    title = "Questions vs Response of Alcohol Consumption",
    xaxis = list(title = "Question",tickfont = list(size = 5)),
    yaxis = list(title = "Response"),
    barmode = "stack"
  )
```

Plot 1 shows the number of individuals who consumed alcohol, resumed drinking alcohol after quitting briefly, or reduced the total number of drinks consumed. The x-axis shows the indicators from the CDC 2007 PRAM data set, and y-axis shows the data values. Overall, it looks like the number of mothers who had the same drinks or more had the highest amount of data values. 

### Plots of Maternal Tobacco Use

```{r}
library(ggplot2)
library(plotly)
library(dplyr)

# Extract data directly from the original data frame
plot_data <- cleaned_tobac_2007 %>%
  group_by(location_abbr, response) %>%
  summarize(count = n())

# Convert data to Plotly
plot_ly(data = plot_data, x = ~location_abbr, y = ~count, color = ~response, type = "bar", split = ~response) %>%
  layout(
    title = " Tobacco Use by State",
    xaxis = list(title = "Question"),
    yaxis = list(title = "Response"),
    barmode = "stack"
  )

```

Plot 2 shows tobacco consumption by state according to the 2007 CDC PRAM data set. The highest data values were amongst those who answered “yes” and the lowest data values were among those who were specific and answered “41+/ day.”


### Map of Maternal Alcohol Use

```{r, message=FALSE, warning=FALSE}
leaflet() |> 
  addTiles() |> 
  addCircleMarkers(data = cleaned_alc_2007,
                   lng = ~longitude,  # Adjust column name if needed
                   lat = ~latitude,   # Adjust column name if needed
                   label = ~location_abbr,   # Assuming 'Group.1' is a column in your data
                   radius = ~data_value * 0.12,
                   color = "blue",
                   stroke = TRUE,
                   fillOpacity = 0.1,
                   popup = ~paste("Response:", response)) 
```

This map visualizes data points from the `cleaned_tobac_2007` dataset, showcasing the prevalence of Tobacco Use during pregnancy across different states. The sizes of the circles represent the values in the `data_value` column, reflecting the number of participants who responded affirmatively to Tobacco Use during pregnancy in each state. Larger circles indicate a higher number of positive responses, offering an overview of the distribution of this behavior across geographic regions
  
  
### Map of Maternal Tobacco use

```{r, message=FALSE, warning=FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = cleaned_tobac_2007,
                   lng = ~longitude,  
                   lat = ~latitude,   
                   label = ~location_abbr,   
                   radius = ~data_value * 0.12,
                   color = "magenta",
                   stroke = TRUE,
                   fillOpacity = 0.1,
                   popup = ~paste("Response:", response),
                   group = ~location_abbr)
```

This map illustrates data points from the `cleaned_tobac_2007` dataset, depicting the prevalence of Tobacco Use during pregnancy across various states. The sizes of the circles correspond to the values in the `data_value` column, reflecting the number of participants who reported affirmative responses to Tobacco Use during pregnancy in each state. Larger circles indicate a higher number of positive responses, providing insight into the regional distribution of this behavior


### Map of Infant Mortality Rate


```{r}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = cleaned_infant_mortality,
                   lng = ~longitude,
                   lat = ~latitude,
                   label = ~location_abbr,
                   radius = ~data_value * 0.12,
                   color = "orange",
                   stroke = TRUE,
                   fillOpacity = 0.5,
                   popup = ~paste("Response:", response))
```

This Leaflet plot provides a visual representation of infant mortality data on a map. The size of the circles corresponds to the values in the `data_value` column, indicating the rate of infant mortality for each location.
  
  
### Infant Mortality by State 

The plot above shows the locations of infant mortality rate across the US. 

```{r}
infant_deaths <- cleaned_infant_mortality %>%
  filter(question == "Indicator of infant currently alive" & response == "NO") %>%
  group_by(location_desc) %>%
  summarize(total_infant_deaths = n())

# Display the table using knitr::kable()
knitr::kable(infant_deaths)

```


The table provides a summary of total infant deaths by state, with each row representing a specific location. The `location_desc` column denotes the state, and the `total_infant_deaths` column indicates the corresponding number of infant deaths in each location. The data suggests variability in infant mortality rates across different regions, with some areas reporting higher or lower rates than others. For instance, states like Pennsylvania have a notably lower count of infant deaths, while others, such as Alaska and Arkansas, have higher counts. However, most of the data seemed to stay within the 35 to 50 range. This summary provides an overview of the distribution of infant deaths across various geographical locations. 

### Looking at infant mortality rates by race, income levels, age,  education levels and intent. 

#### Table showing Infant Mortality by Race/Ethnicity

```{r}
filtered_mortality_race <- cleaned_infant_mortality %>%
  filter(break_out_category == "Maternal Race/Ethnicity" &
         (break_out %in% c("Hispanic", "Non-hispanic", "White, non-Hispanic")) &
         question == "Indicator of infant currently alive" & response == "NO")

# Display the table using knitr::kable()
knitr::kable(filtered_mortality_race)

view(filtered_mortality_race)
```

This table shows in depth information about infant mortality rate by race-ethnicity in the 31 states in this sample. The race/ethnicity variable was re-categorized into just "Hispanic" and "White-non Hispanic" to give us our variables of interest. Using this information, we will visualize the relationship between infant mortality rate and how race/ethnicity plays a role. 


#### Plot showing the effect of Race/Ethnicity on Infant Mortality 

```{r}
plot_infant_deaths <- ggplot(filtered_mortality_race, aes(x = break_out, fill = break_out)) +
  geom_bar() +
  labs(title = "Infant Deaths by Ethnicity",
       x = "Ethnicity",
       y = "Total Infant Deaths") +
  scale_fill_manual(values = c("Hispanic" = "blue", "Non-hispanic" = "green", "White, non-Hispanic" = "pink")) +
  theme_minimal()

print(plot_infant_deaths)
```

The `plot_infant_deaths` above shows a plot of infant deaths categorized by whether they were Hispanic or not. The graph shows that those who were not Hispanic had a higher infant death count than those who were Hispanic.


#### Plot of Infant Mortality and Maternal Income

```{r}

filtered_mortality_income <- cleaned_infant_mortality %>%
  filter(break_out_category == "Income (years 2004 and beyond)" &
         (break_out %in% c("Less than $10,000", "$10,000 to $24,999", "$25,000 to $49,999", "$50,000 or more")) &
         question == "Indicator of infant currently alive" & response == "NO")

view(filtered_mortality_income)

plot_infant_income <- ggplot(filtered_mortality_income, aes(x = break_out, fill = break_out)) +
  geom_bar() +
  labs(title = "Infant Deaths by Income",
       x = "Maternal Income",
       y = "Total Infant Deaths") +
  scale_fill_manual(values = c("Less than $10,000" = "blue", "$10,000 to $24,999" = "red", "$25,000 to $49,999" = "purple", "$50,000 or more" = "pink")) +
  theme_minimal()

print(plot_infant_income)

```

This plot shows the relationship between maternal income levels and infant mortality rate. From the bar graph, those in the income bracket of $25,000-$49,000 had a higher infant mortality rate and those who made more than $50,000 had the lowest. This was an interesting observation, as those who made less that $10,000 a year would be expected to have the highest infant mortality as low income levels is a social determinant of infant health outcomes. 


#### Plot of Infant Mortality and Maternal Age 

```{r}
filtered_mortality_age <- cleaned_infant_mortality %>%
  filter(break_out_category == "Maternal Age - 18 to 44 years in groupings" &
         (break_out %in% c("Age < 18", "Age 18 - 24", "Age 25 - 29", "Age 30 - 44", "Age 45+")) &
         question == "Indicator of infant currently alive" & response == "NO")

plot_mortality_age <- ggplot(filtered_mortality_age, aes(x = break_out, fill = break_out)) +
  geom_bar() +
  labs(title = "Infant Deaths by Maternal Age",
       x = "Maternal Age",
       y = "Total Infant Deaths") +
  scale_fill_manual(values = c("Age < 18" = "blue", "Age 18 - 24" = "purple", "Age 25 - 29" = "pink", "Age 30 - 44" = "yellow", "Age 45+" = "orange")) +
  theme_minimal()

print(plot_mortality_age)

```


This plot shows infant mortality rates by age. Interestingly, maternal populations below the age of 18 have the lowest infant mortality rates. Comparatively, women in the 45+ age category have a higher infant mortality rate. Women in the 18-24, 25-29 and 30-44 all have approximately the same levels of infant mortality rates. 


#### Plot of Maternal Education and Infant Death 

```{r}
filtered_mortality_educ <- cleaned_infant_mortality %>%
  filter(break_out_category == "Maternal Education" &
         (break_out %in% c("<12 yrs", "12 yrs", ">12 yrs")) &
         question == "Indicator of infant currently alive" & response == "NO")

plot_mortality_educ <- filtered_mortality_educ |>
ggplot(aes(x = break_out, fill = break_out)) +
  geom_bar() +
  labs(title = "Infant Deaths by Maternal Education",
       x = "Maternal Education",
       y = "Total Infant Deaths") +
  scale_fill_manual(values = c("<12 yrs" = "blue", "12 yrs" = "purple", ">12 yrs" = "pink")) +
  theme_minimal()

print(plot_mortality_educ)
```

This plot shows the relationship between maternal education and infant mortality rates. This is an interesting finding that shows those with greater than 12 years of education had higher infant mortality rates compared to those with less than 12 years of education. This could be because factors like race could have been a confounding variable. Additionally, it can also be assumed that people who were in school longer were also older, so that could also be an additional factor. 

#### Plot of Maternal Medicaid Recepient and Infant Death

```{r}
filtered_mortality_medi <- cleaned_infant_mortality %>%
  filter(break_out_category == "Medicaid Recipient" &
         (break_out %in% c("Non-Medicaid", "Medicaid")) &
         question == "Indicator of infant currently alive" & response == "NO")

plot_mortality_medi <- filtered_mortality_medi |>
ggplot(aes(x = break_out, fill = break_out)) +
  geom_bar() +
  labs(title = "Infant Deaths by Medicaid Recpient",
       x = "Medicaid Recepient",
       y = "Total Infant Deaths") +
  scale_fill_manual(values = c("Non-Medicaid" = "blue", "Medicaid" = "purple")) +
  theme_minimal()

print(plot_mortality_medi)
```

This plot shows the relationship between infant mortality rate and mothers who were recipients of medicaid. This bar graph shows that there is no statistical difference. 


#### Plot of Maternal Intent and Infant Death

```{r}
filtered_mortality_intent <- cleaned_infant_mortality %>%
  filter(break_out_category == "Pregnancy Intendedness" &
         (break_out %in% c("Unintended", "Intended")) &
         question == "Indicator of infant currently alive" & response == "NO")

plot_mortality_intent <- filtered_mortality_intent |>
ggplot(aes(x = break_out, fill = break_out)) +
  geom_bar() +
  labs(title = "Infant Deaths by Maternal Intent",
       x = "Maternal Intent",
       y = "Total Infant Deaths") +
  scale_fill_manual(values = c("Unintended" = "orange", "Intended" = "red")) +
  theme_minimal()

print(plot_mortality_intent)
```

This plot shows the relationship between maternal intent to have children with infant mortality rate. Those who intended to have a child, actually had a higher rate of infant mortality rate. There could be several confounding factors that could contribute to this, such as age and stress factors. 
